{\TheTitle}\thanks{Submitted to the editors DATE. \funding{This work was funded by the Fog Research Institute under contract no.~FRI-454.}}

### 2017-10-17 ssc-0-0.tex
complete a first draft of the semi-implicit paper

\usepackage{amsopn}
\DeclareMathOperator{\diag}{diag}

@@

% formula equations
The semi-implicit formula is a rule for integrating systems of
ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
is often used as a method of lines, to solve stochastic partial
differential equations that occur when phase space Monte-Carlo
methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
These equations are related to Schr\"odinger's equation,
\begin{equation}
    \frac{\partial\psi}{\partial t} = -\del^2\psi + V\psi,
\end{equation}
with additional nonlinear and random terms.  The spatial axes are
discretised, often using spectral derivative operators, to give a
system of ordinary or stochastic differential equations
\cite{2000-Trefethen-Spectral}, and the semi-implicit formula is
used to integrate these equations.

{\bf Relate Hilbert space expansions to methods of lines using sync wave packets.}

For a system of equations~${dx\over dt}=f(x,t)$, with time step~$h$,
the formula is
\begin{equation}\label{eq:i}
    x_{n+1}=x_n+hf_{n+\frac12},
\end{equation}
where
\begin{equation}\label{eq:ii}
    f_{n+\frac12}=f\left({x_n+x_{n+1}\over 2}\right).
\end{equation}

The formula has the theoretical advantage that it conserves Hilbert space norm when it is applied to a quantum equation of motion of the form
\begin{equation}\label{eq:iii}
	{dx\over dt}=-iHx,
\end{equation}
where $H$ is a hermitian operator.  However, it is mostly used because it has been found to work well in practice.

It is not obvious how to apply this formula, because of the need
to evaluate~$x_{n+1}$ and~$f_{n+\frac12}$ self-consistently.  The
usual way of doing this is a very simple iterative one.  At first,
$x_{n+1}$ is set to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then
Equations~\cref{eq:i,eq:ii} are evaluated in turn, and the values of~$x_n$
and~$f_{n+\frac12}$ updated.  This ought to continue until they
converge, but in practice this is continued for a fixed number of
iterations.  Formally, this process gives a sequence of~$x_{n+1}^j$
and~$f_{n+\frac12}^j$, with
\begin{subequations}
\begin{align}
	\label{eq:iv} x_{n+1}^0&=x_n\qquad f_{n+\frac12}^0=f(x_n)\\
	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}=f\left({x_n+x_{n+1}^j\over 2}\right),\qquad j\ge0.
\end{align}
\end{subequations}
Any fixed point of this iteration is a self-consistent solution to
Equations~\cref{eq:i,eq:ii}, but the question remains of when the
iteration converges.

It is a generally known problem that the time steps in formulae for
partial differential equations need to be reduced in proportion to
the spatial grid steps, in order for the formulae to remain stable.
To date, no one has made a specific analysis of the iterative
semi-implicit formula, so the stable time step has had to be
determined by trial and error.  This can be a time consuming process
for fields that are being simulated on large grids.

@@

Suppose that the equation being integrated has the form of
Equation~\cref{eq:iii}, where~$x$ is a vector of coefficients
obtained by expanding a ket~$|\psi\rangle$ over an orthonormal
basis, and~$f$ is the linear operator~$-iH$, where~$H$ is the matrix
for the Hamiltonian operator in this basis.  The
iteration rules of Equation~\cref{eq:v} become
$$x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j$$
or
$$x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j$$
with solution
$$x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.$$
The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$ as the
hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the sums of a geometric series in~$\omega_m$, with the relation between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$\omega_m^j$ of~$U_j$.  The lightest line is for~$j=1$, and they get heavier for~$j=2$, 3, 4.  The exact time evolution operator~$U$ is unitary, with unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The sequence of~$U_j$ is a power series, which will converge if and only if the angular frequencies~$\omega_m$ are bounded by~$2/h$, a kind of Nyquist
condition.  If the discretised Hamiltonian has an eigenstate~$x_\omega$
with a frequency~$\omega\gg 2/h$, then~$U$ will magnify the component
of~$x_\omega$ in~$x$ by a factor close to~$(h\omega/2)^i$, and this
component is a parasitic solution growing steadily at the
rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.

The semi-implicit method is rarely used to solve the linear Equation~\cref{eq:iii}, and more often used to solve a nonlinear Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
\begin{equation}
	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
\end{equation}
This linearisation has been extensively studied, and the operator~$\hat
H'$ is the hamiltonian for Bogoliubov sound waves \cite{something}.
The stability conditon for the Gross-Pitaevskii~equation is that
the highest frequency of the Bogoliubov sound waves in the discretised
Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
high-frequency Bogoliubov sound waves have the same frequencies as
the eigenstates of~$\hat H$.  Most simulations use grids fine enough
for this to be the case, and the nonlinear term will not affect the
stability of such simulations.

@@

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
  \label{fig:i}
\end{figure}

@@

The semi-implicit formula considered here is a special case, but most ODE formulae will become unstable when frequencies get large compared to time steps.

Stiff formulae stay stable, but reject the physics that is happening on fast timescales.  If we want to simulate physics at frequency~$\omega$, it seems reasonable to require time steps~$?$.

The problem comes when the interesting physics occurs at slow timescales, but the state representation is inadvertently capable of representing high energies that evolve on fast timescales.

In the case here, where we expanded over energy eigenstates, you can simply truncate.  This is contrived: if you know the energy eigenstates, you already know the answer, and you don't need to simulate anything.

In general, you will use a variational ansatz, or some other tricky representation.  In this case, some care must be taken to avoid inadvertently representing high energy states.  Details to come in future papers.

\section*{Acknowledgments}
We would like to acknowledge the assistance of volunteers in putting
together this example manuscript and supplement.

>>>>

\usepackage{amsopn}
\DeclareMathOperator{\sinc}{sinc}

@@

% formula equations
The semi-implicit formula is a rule for integrating systems of
ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
is often used as a method of lines, to solve stochastic partial
differential equations that occur when phase space Monte-Carlo
methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
These equations are related to Schr\"odinger's equation,
\begin{equation}\label{eq:vi}
    \frac{\partial\psi}{\partial t} = -i(-\nabla^2 + V)\psi,
\end{equation}
with additional nonlinear and random terms.  The spatial axes are
discretised, often using spectral derivative operators, to give a
system of ordinary or stochastic differential equations
\cite{2000-Trefethen-Spectral}.  Then these equations are integrated
by the semi-implicit formula.

For a system of equations~${dx\over dt}=f(x,t)$,
the semi-implicit formula with time step~$h$ is
\begin{equation}\label{eq:i}
    x_{n+1}=x_n+hf_{n+\frac12},
\end{equation}
where
\begin{equation}\label{eq:ii}
    f_{n+\frac12}=f\left({\textstyle\frac{x_n+x_{n+1}}{2}}\right).
\end{equation}
When this formula is applied to equations of a form similar to~\cref{eq:vi},
\begin{equation}\label{eq:iii}
	{dx\over dt}=-iHx,
\end{equation}
where~$H$ is a hermitian operator acting on the complex vector~$x$,
there is an advantage in principle because the norm of~$x$ is
conserved in the discretised solution.  {\bf Expand on this.}  However, the formula is
used mainly because it has been found to work well in practice.

% iterations
It is not obvious how to implement this formula, because~$x_{n+1}$
and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
usual way is a simple iteration.  To start,
$x_{n+1}$~is initialised to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then each iteration
evaluates Equations~\cref{eq:i,eq:ii} in turn, and updates the values
of~$x_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the iterations generate a sequence of~$x_{n+1}^j$
and~$f_{n+\frac12}^j$, with
\begin{subequations}
\begin{align}
	\label{eq:iv} x_{n+1}^0&=x_n\qquad f_{n+\frac12}^0=f(x_n)\\
	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j
		\qquad f_{n+\frac12}^{j+1}=f\left({\textstyle\frac{x_n+x_{n+1}^j}{2}}\right),\qquad j\ge0.
\end{align}
\end{subequations}
This continues until the sequences converge, which is usually
achieved after a small, fixed number of iterations.  Any fixed point
of the iteration is a self-consistent solution to
Equations~\cref{eq:i,eq:ii}.  The remaining question is under what
conditions the iterations converge.

% steps stability
Generally, when a partial differential equation is discretised by
a given formula, the time step needs to be reduced in proportion
to the grid spacing in order for the discretised solution to remain
stable \cite{something}.  No one has analysed the stability of the
semi-implicit formula with iterations.  The stable time step has
had to be determined by trial and error, a time consuming process
for three-dimensional wave functions that are being simulated on
large grids.

@@

% basis coefficients
When the semi-implicit rule is applied to \cref{eq:vi}, or a
stochastic or nonlinear variant, the laplacian operator is usually
evaluated by a spectral method.  This is equivalent to expanding
the wave function~$\psi(q)$ over a basis of sinc wave packets
\begin{equation}
	\psi(q) = \sum x_i\phi_i(q),
\end{equation}
{\bf figure out range of $i$}
where
\begin{equation}
	\phi_i(q) = \frac1{\sqrt{l}} \sinc(\textstyle\frac{q}{l} - i).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc{q} = \frac{\sin(\pi q)}{\pi q},
\end{equation}
under which the functions~$\phi_i$ are an orthonormal basis for the
space of functions limited to bandwidth~$1/2l$.  The spectral method
reduces \cref{eq:vi} to an equation of the form~\cref{eq:iii} for
the coefficient vector~$x$, where
\begin{equation}
	H_{ij} = \int \phi_i(q) (-\nabla^2 + V(q)) \phi_j^\ast(q).
\end{equation}

At this point, the
iteration rules of Equation~\cref{eq:v} become
\begin{equation}
x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j
\end{equation}
or
\begin{equation}
x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j
\end{equation}
with solution {\bf check factor of 2}
\begin{equation}\label{eq:viii}
x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.
\end{equation}
The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$
as the hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the
sums of a geometric series in~$\omega_m$, with the relation
between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$h\omega_m^j$ of~$U_j$.  The lightest line is for~$j=1$, and they get heavier for~$j=2$, 3, 4.  The exact time evolution operator~$U$ is unitary, with unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The sequence of~$U_j$ is a power series, which will converge if and
only if the angular frequencies~$\omega_m$ are bounded by the reciprocal time step~$2/h$, a
kind of Nyquist condition.  If the discretised hamiltonian has an
eigenstate~$x_\omega$ with a frequency~$\omega\gg 2/h$, then~$U$
will magnify the component of~$x_\omega$ in~$x$ by a factor close
to~$2(h\omega/2)^i$, and this component is a parasitic solution
growing steadily at the rate~$\frac{\ln2}h\frac{i}{h}\ln(\frac{h\omega}2)$.

{\bf Rewrite for clarity.}
The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
\begin{equation}
	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
\end{equation}
This linearisation has been extensively studied, and the operator~$\hat
H'$ is the hamiltonian for Bogoliubov sound waves \cite{something}.
The stability conditon for the Gross-Pitaevskii~equation is that
the highest frequency of the Bogoliubov sound waves in the discretised
Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
high-frequency Bogoliubov sound waves have the same frequencies as
the eigenstates of~$\hat H$.  Most simulations use grids fine enough
for this to be the case, and the nonlinear term will not affect the
stability of such simulations.

@@

As a test of this analysis, a quantum Kerr oscillator was simulated
using the semi-implicit method.  This system is a single-orbital
analog of the fields for repulsive particles that are simulated in
quantum optics.  The states with definite numbers of particles are
an orthonormal basis for the quantum Hilbert space, so the coefficient
vector~$x$ has indices~$n=0$, 1, \dots, $N$, where the number of
particles is limited to~$N$ to make the simulation finite.  (In the
experiment shown, $N=25$.)  The hamiltonian matrix is then diagonal,
with
\begin{equation}
	H_{nn} = \omega_n = n(n-9)/2.
\end{equation}
The eigenvalues range from~$\omega_4=\omega_5=-10$,
through~$\omega_1=\omega_9=0$, up to~$\omega_{25}=200$.

In terms of wave functions and Schr\"odinger's equation, this could
be regarded as the hamiltonian for a single particle, where~$x_n$
was the coefficient of a Lagrange polynomial~$p_n$ (not quite) in
the wave function.  It would be quite a wierd hamiltonian physically,
with terms in products of position and momentum.  On the other hand,
it makes physical sense in second quantisation, where it describes
a collection of particles occupying the same orbital, with mutual
repulsion.

The oscillator was initialised in a quantum-optical coherent state
\cite{glauber}, where the number of particles is Poisson distributed,
with
\begin{equation}\label{eq:vii}
	x_n = e^{-|\alpha|^2/2}\frac{\alpha^n}{\sqrt{n!}}.
\end{equation}
The experiment was run with~$\alpha=2$, which provides some small
initial coefficients with large eigenfrequencies, which are useful
for demonstrating the growth of the parasitic solutions.

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.  (Tidy up fonts.)}
  \label{fig:i}
\end{figure}

The results of solving this Hamiltonian with the semi-implicit
method, with 4~iterations and time step~$h=0.03$, are shown in
\cref{fig:a}.  The hamiltonian matrix is diagonal, so the norms~$|x_n|$
are constant in the exact solution.  The right hand side of
\cref{eq:vii} is shown as the solid lines at~$t=0$ and~$t=0.09$,
and the values of~$|x_n|$ computed by the semi-implicit formula are
shown as the rulings between these lines.  The size of~$x_n$ is
plotted for all four iterations of the formula within each timestep.
\Cref{eq:viii} shows that, for~$\omega_n\gg 2/h$, the growth of~$x_n$
will still be geometric, resulting in straight lines on the logarithmic
scale of the graph.

Parasitic growth can be
seen at the left hand side, where~$\omega_n$ is large.  The limit
of stability is~$\omega_n=2/h\approx 67$, and the dotted line
indicates~$\omega_17=68$.  The dashed line shows the result of
growing~$|x_n|$ at the predicted rate~$\frac4h\ln(\frac{h\omega}2)$.
The undulations in the lines~$|x_n(t)|$ correspond to the growth
rates given by the different lines in \cref{fig:ii}, which apply
for the fraction of each step ending with a given iteration.

@@

The semi-implicit formula discussed here is only one of the formulae
that can be used to integrate ordinary differential equations, and
that could be applied to solve \cref{eq:iii} numerically.  However,
most of these formulae suffer from similar problems, and become
unstable when the eigenvalues of the hamiltonian matrix are large
compared to the time step.  The phenomenon of stiffness in discretised
partial differential equations is well known \cite{something}.

There are special formulae to integrate stiff differential equations,
which remain stable when the time steps are long compared to the
eigenvalues of~$H$.  However, these remain stable by ignoring any
physics that is happening on the fast timescales corresponding to
the large eigenvalues.  If the goal is to simulate physics at a
given frequency, then it seems reasonable that the simulation will
require a time step around the corresponding Nyquist rate.

The situation where stiff integration formulae would be appropriate
is when the physics being simulated occurs at slow timescales, but
the method by which the physical state is being represented
inadvertently allows the state to contain a component with a large
energy, that evolves on a fast timescale.  In the numerical example,
this occured due to the coefficient vector~$x$ including coefficients
with large~$n$, for which~$x_n$ was too small to have any effect
on the state.  As demonstrated, the formula caused these minute
components to grow quickly and dominate the actual solution.  It would
be possible to address this by using an integration formula that
was adapted to stiff differential equations.  But it might be better
to address this by changing the representation so that the simulated
state can not include such components, or to modify the hamiltonian
dynamics so that they evolved with reasonable frequencies in the
simulation.

In the example here, the representation was simply an expansion
over energy eigenstates.  This can be modified very simply, by
reducing~$N$ to exclude the large eigenvalues and frequencies.  In
the example, an expansion with coefficients~$x_0$ to~$x_{17}$ would
be stable with the time step used in the example.  However, this
is a contrived example.  If the energy eigenstates were known, their
amplitudes could be calculated directly as complex
exponentials~$e^{i\omega_nt}$, and there would be no need to use a
potentially unstable integration formula.

In real simulations, quantum states are represented by expansion
over a set of states that are not energy eigenstates, by use of a
variational ansatz, or by similar methods.  These representations
are not directly related to the hamiltonian, so it is often the
case that some states with spuriously high energies can be represented,
and the resulting ordinary differential equations are unnecessarily
stiff.  In these simulations, some care must be taken to avoid
inadvertently representing high energy states, or to modify the
hamiltonian dynamics to prevent them from causing stiffness.  Details
for one representation, using Gaussian wave packets, will come in
future papers.

\section*{Acknowledgments}
I would like to acknowledge the support of Swinburne University of Technology.
